Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages

### 2013-08-14 paper/main.tex
Update introduction

In this paper, we apply this technique to web programming, and introduce an EDSL to facilitate the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks such as SQL injection. We also show how EDSLs may be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling. 
\end{abstract}

@@

\section{Introduction}
Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits such as SQL injection \cite{owasp:sqli} or cross-site scripting \cite{owasp:xss}. Security breaches using such exploits are far-reaching, and high profile cases involve large corporations such as Sony, who suffered a well-publicised and extremely costly SQL Injection breach in 2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012 \cite{imperva:yahoo}.

@@

%%------- EXAMPLE
Allow us to consider a simple, traditional database access routine, written in PHP, where we wish to obtain the name and address of every employee working in a given department, \texttt{\$dept}. We firstly construct an object representing a database connection, where the arguments are the database host, user, password and name respectively:
%
\begin{Verbatim}
$conn = new mysqli("localhost", "username", 
                    "password", "db");
\end{Verbatim}
%
We must subsequently check to see if the connection was successful, and cease execution of the database routine if so. Of course, this check is optional, so it would be perfectly possible to omit the check: this would, however, cause problems with later execution steps.
%
\begin{Verbatim}
  if (mysqli_connect_errno()) {
      exit();
  }
\end{Verbatim}
%
We then create a prepared statement detailing our query, and bind the \texttt{`dept'} value:
\begin{Verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param('s', $dept);
\end{Verbatim}
%
After the parameters have been bound, we must execute the statement, assign variables into which results will be stored, and fetch each row in turn. At this point, a developer may forget to execute the statement before beginning to fetch results or may have forgotten to bind the parameters to the statement:
%
\begin{Verbatim}
  $stmt->execute();
  $stmt->bind_result($name, $address);
  while ($stmt->fetch()) {
    printf("Name: %s, Age: %s", $name, $age);
  }
\end{Verbatim}
%
Finally, once the statement and connection are no longer needed, they should be closed in order to discard the associated resources:
%
\begin{Verbatim}
  $stmt->close();
  $conn->close();
\end{Verbatim}
%
Even in this small example, there exists a specific resource usage protocol. Firstly, a connection to the database must be opened. The object-oriented style used in the example encapsulates this to an extent, as the object must be created in order for operations to be performed, however it is less obvious in a procedural version of the code. Secondly, a prepared statement is created, using the raw SQL and placeholders to which variables are later bound. The statement is then executed, and each row is retrieved from the database. Finally, the resources are freed.

There is, however, no way to statically guarantee that this protocol is adhered to. A developer may, for example, accidentally close a statement whilst still retrieving rows, which would cause a runtime error. Similarly, a programmer may omit closing the statement or connection, which can lead to substantial problems such as handle leaks in longer-running server applications.

An alternative approach is to ensure that a program is semantically sound before it is run. Through the use of \textit{dependent types}, it is possible to impose arbitrarily strong specifications on programs, requiring that these specifications are fulfilled prior to being accepted by a type checker. A large drawback of this, however, is that automatic verification by the compiler is often impossible, requiring additional proofs to be specified by the developer. This severely limits the uptake of dependently-typed languages in real-world settings.

This complexity can be ameliorated through the use of \textit{embedded domain-specific languages} (EDSLs), which aim to abstract away from the complexity of the underlying dependent type mechanisms, allowing users to write verfied domain-specific code without having to consider the underlying verification mechanisms.   %TODO: rephrase this

\idris{} \cite{brady2011idris} is a language with full dependent types, and extensive support for embedded domain-specific languages through ad-hoc syntax overloading and syntax macros. Through the use of \idris{}, and the \texttt{Effects} framework \cite{brady:effects}, we present a dependently-typed web framework, which allows for the construction of programs with additional guarantees about correctness and security, whilst minimising the increase in development complexity.

\subsection{Contributions}
The primary contribution of this paper is the application of previous work to show that dependent types may be used to provide stronger static guarantees about the correctness and security of web applications, whilst minimising additional development complexity. In particular:

\begin{itemize}
\item A form-handling mechanism, which preserves type information and therefore increases applications' resilience to attacks such as SQL injection.
\item Representations of CGI, SQLite and a simple session handler as \textit{algebraic effects}, allowing programs to be accepted only when they adhere to specific resource usage protocols.
\item A message board application making use of the framework, and showing that the framework may be used for practical applications.
\end{itemize}

@@

An effect is represented as an instance of the \texttt{EFFECT} type. Such instances may be associated with a \textit{resource}, 
Describing a program in the \texttt{Effects} library is achieved using the following data type:
\begin{Verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) ->
       (es’ : List EFFECT) ->
       (a : Type) -> Type
\end{Verbatim}
\texttt{EffM} is a type that is parameterised over an execution context (\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list of effects at the end of the program's execution (\texttt{es'}), and a return type (\texttt{a}). The list of input types can be used to ensure that an operation may only be called when the program is in a particular state, and since the list of input effects may differ from the list of output effects, we may modify the resource associated with the state after each operation. This allows us to simply enforce a resource usage protocol, and deviation from this protocol will result in a compile-time error.

Implementing a new effect involves creating a new algebraic data type, of kind \texttt{Effect}, which describes the available operations and how they modify the resource associated with the effect. Each operation consists of an initial resource \texttt{res}, the resource at the end of the operation \texttt{res'}, and the output type of the operation \texttt{t}.
%TODO: Execution
\begin{Verbatim}
Effect : Type
Effect = (res : Type) -> (res’ : Type) ->
         (t : Type) -> Type
\end{Verbatim}
Operations are made available to users by writing functions with the corresponding \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
\begin{Verbatim}
queryPostVar : 
 String -> 
 Eff m [CGI (InitialisedCGI TaskRunning)] 
   (Maybe String)
queryPostVar query = (QueryPostVar query)
\end{Verbatim}
Additionally, we define \textit{handlers}, which describe how each operation is interpreted in a given execution context. Finally, \texttt{Effect} instances are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff} constructor, which is provided with the initial resource value.

>>>>

In this paper, we apply this technique to web programming, and introduce an
EDSL to facilitate the creation and handling of web forms which retain type
information, reducing the scope for programmer error and attacks such as SQL
injection. We also show how EDSLs may be constructed to enforce resource usage
protocols associated with common web operations such as CGI, database access
and session handling.

\end{abstract}

@@

\section{Introduction}

Web applications, whilst ubiquitous, are also prone to incorrect construction
and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
scripting \cite{owasp:xss}. Security breaches using such exploits are
far-reaching, and high profile cases involve large corporations such as Sony,
who suffered a well-publicised and extremely costly SQL Injection breach in
2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
\cite{imperva:yahoo}.

@@

%%------- EXAMPLE
Let us consider a simple database access routine, written in
PHP, where we wish to obtain the name and address of every employee working in
a given department, \texttt{\$dept}. We firstly construct an object
representing a database connection, where the arguments are the database host,
user, password and name respectively:

%
\begin{Verbatim}
$conn = new mysqli("localhost", "username", 
                    "password", "db");
\end{Verbatim}
%
We then check to see if the connection was successful, and exit
if not.  This check is optional, so
it would be possible to omit it: this would, however, cause
problems with later steps.

%
\begin{Verbatim}
  if (mysqli_connect_errno()) {
      exit();
  }
\end{Verbatim}
%
We then create a prepared statement detailing our query, and bind the
\texttt{`dept'} value:

\begin{Verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param('s', $dept);
\end{Verbatim}
%
After the parameters have been bound, we execute the statement, assign
variables into which results will be stored, and fetch each row in turn. At
this point, a developer may forget to execute the statement before beginning to
fetch results or may have forgotten to bind the parameters to the statement:

%
\begin{Verbatim}
  $stmt->execute();
  $stmt->bind_result($name, $address);
  while ($stmt->fetch()) {
    printf("Name: %s, Age: %s", $name, $age);
  }
\end{Verbatim}
%
Finally, once the statement and connection are no longer needed, they should be
closed in order to discard the associated resources:

%
\begin{Verbatim}
  $stmt->close();
  $conn->close();
\end{Verbatim}
%
Even in this small example, there exists a specific resource usage protocol.
Firstly, a connection to the database must be opened. The object-oriented style
used in the example encapsulates this to an extent, as the object must be
created in order for operations to be performed, however it is less obvious in
a procedural version of the code. Secondly, a prepared statement is created,
using the raw SQL and placeholders to which variables are later bound. The
statement is then executed, and each row is retrieved from the database.
Finally, the resources are freed.

Problems may arise if the protocol is not followed correctly.
A developer may, for example, accidentally close a statement whilst still
retrieving rows, which would cause a runtime error. Similarly, a programmer may
omit closing the statement or connection, which can lead to
problems such as resource leaks in longer-running server applications.
However, in conventional programming languages, there is no way to check
automatically that a protocol is followed.

In contrast, the use of \textit{dependent types} makes it possible
to specify a program's behaviour precisely, and to check that a 
specification is followed.
%
The difficulty is 
that automatic verification by a compiler can be difficult or
often impossible, requiring additional proofs to be given by the programmer.

This complexity can be ameliorated through the use of \textit{embedded
domain-specific languages} (EDSLs), which aim to abstract away the
complexity of the underlying type system. EDSLs allow domain experts to
focus on a problem domain, rather than the specific details of their
solution.

\idris{} \cite{brady2011idris} is a language with full dependent types, and
extensive support for EDSLs through overloading and syntax macros. Through the
use of \idris{}, and a framework for describing resource protocols using
\emph{algebraic effects}~\cite{brady:effects}, we
present a dependently-typed web framework, which allows the construction of
programs with additional guarantees about correctness and security, whilst
minimising the increase in development complexity.

\subsection{Contributions}
The primary contribution of this paper is the application of 
dependent types may to provide strong static guarantees
about the correctness and security of web applications, whilst minimising
additional development complexity. In particular, we present:

\begin{itemize}
\item A form-handling mechanism, which preserves type information and
manages user input, therefore
increasing applications' resilience to attacks such as SQL injection.

\item Representations of CGI, Databases and sessions as
\textit{algebraic effects}, allowing programs to be accepted only when they
follow clearly defined resource usage protocols.

\item A practical demonstration of the framework, a message board application.

\end{itemize}

@@

An effect is represented as an instance of the \texttt{EFFECT} type. Such
instances may be associated with a \textit{resource}, Describing a program in
the \texttt{Effects} library is achieved using the following data type:

\begin{Verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) ->
       (es’ : List EFFECT) ->
       (a : Type) -> Type
\end{Verbatim}
\texttt{EffM} is a type that is parameterised over an execution context
(\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list
of effects at the end of the program's execution (\texttt{es'}), and a return
type (\texttt{a}). The list of input types can be used to ensure that an
operation may only be called when the program is in a particular state, and
since the list of input effects may differ from the list of output effects, we
may modify the resource associated with the state after each operation. This
allows us to simply enforce a resource usage protocol, and deviation from this
protocol will result in a compile-time error.

Implementing a new effect involves creating a new algebraic data type, of kind
\texttt{Effect}, which describes the available operations and how they modify
the resource associated with the effect. Each operation consists of an initial
resource \texttt{res}, the resource at the end of the operation \texttt{res'},
and the output type of the operation \texttt{t}.

%TODO: Execution
\begin{Verbatim}
Effect : Type
Effect = (res : Type) -> (res’ : Type) ->
         (t : Type) -> Type
\end{Verbatim}

Operations are made available to users by writing functions with the
corresponding \texttt{EffM} type, which may then be used to invoke the
operation from within an effectful program.

\begin{Verbatim}
queryPostVar : 
 String -> 
 Eff m [CGI (InitialisedCGI TaskRunning)] 
   (Maybe String)
queryPostVar query = (QueryPostVar query)
\end{Verbatim}

Additionally, we define \textit{handlers}, which describe how each operation is
interpreted in a given execution context. Finally, \texttt{Effect} instances
are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff}
constructor, which is provided with the initial resource value.
